RMSEA lower bound: want less than 0.05

RMSEA upper bound: want less than 0.10

SRMR: want less than 0.08


# Factor analysis

```{r}
library(foreign)
library(lavaan)
library(lmSupport)
library(psych)
d <- lm.readDat("data/data_apgar.dat")

m1 <- '
  apgar ~ anninc
  gestat ~ anninc
  smokes ~ anninc
  prenat ~ anninc
  wgtgain ~ anninc
  gender ~ anninc
'

# For some reason lavaan estimates covariances of endogenous variables

fit1 <- sem(m1, data = d, likelihood = "wishart")
summary(fit1, standardized = TRUE, fit.measures = TRUE)

```














# Confirmatory factor analysis

We want to see if a latent variable predicts these variables. The latent variable needs a scale. We set `Fsingle` to have the same scale as `visperc`. 

* Endogenous variables are now indicators
* Distrurbances are now measurement errors.
* We still estimate the variance of `Fsingle`

## Degrees of Freedom

```
p = 21

q = 0 causal paths + 
    0 covariances of exogenous variables + 
    5 factor loadings (1 is fixed to one) + 
    1 variances of latent factor + 
    6 measurement errors 
q = 12 

df = 9
```

```{r}
data_intell <- read.spss("data/Grnt_fem.sav", to.data.frame = TRUE)
describe(data_intell)

m1 <- '
  # Latent variable definition.  
  Fsingle =~ visperc + cubes + lozenges + paragrap + sentence + wordmean 
  # Fsingle causes these things on the right
'

m2 <- '
  # Latent variable definition.  
  Fsingle =~ cubes + lozenges + paragrap + sentence + wordmean + visperc
  # Fsingle causes these things on the right
'

fit1 <- cfa(m1, data = data_intell, likelihood = "wishart")
fit2 <- cfa(m2, data = data_intell, likelihood = "wishart")

summary(fit1, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
inspect(fit1, "sampstat")
fitted(fit1)
resid(fit1)
resid(fit1, type = "cor")
resid(fit1, type = "standardized")

summary(fit2, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)


```

This is a very poor model.

Don't get a test of estimate for the path we fixed to one.

We say "visual perception loads on to the factor".

The estimates change if the scale of the latent factor change.





## Two factor model

```
p = 21
q = 0 + 1 + 4 + 2 + 6 = 13
```

```{r}


m3 <- '
  spa =~ visperc + cubes + lozenges
  ver =~  paragrap + sentence + wordmean
'

fit3 <- cfa(m3, data = data_intell, likelihood = "wishart")

summary(fit3, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
fitted(fit3)
resid(fit3)
resid(fit3, type = "cor")
resid(fit3, type = "standardized")

source("power.r")
nrow(data_intell)
not_close_fit(df=8, n=73)
close_fit(df=8, n=73)
```








