Lecture 11/14/13 - Structural Regression
===============================================================================

Make a structural regression model.

```{r, tidy = FALSE, warning = FALSE, message = FALSE}
options(width = 100)
library(lavaan)
library(foreign)
library(semPlot)
d <- read.spss("../data/data_carnet3.sav", to.data.frame = TRUE)
# Keep just the `mpn` columns
d <- d[, paste("mpn", 1:10, sep = "")]
m <- "
  F1 =~ mpn2 + mpn4 + mpn6 + mpn9
  F2 =~ mpn1 + mpn7 + mpn10
  F3 =~ mpn3 + mpn5 + mpn8
  F2 ~ F3 + F1
  F3 ~ F1"
fit <- sem(m, d, likelihood = "wishart")
summary(fit, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

This model has the same chi-square as the `m2` model from above; it is an equivalent model. We basically turn covariance paths to causal paths. We need to check whether the structural part of the model (the paths among the factors) is just-identified or over-identified. The other part of the model is the measurement component of the model.

If we fix one path to zero in the structural part of the model, the chi-square of the model will increase. 

If you start out with a poor measurement model, your structural model can only be worse.

```{r, tidy = FALSE}
m4 <- "
  F1 =~ mpn2 + mpn4 + mpn6 + mpn9
  F2 =~ mpn1 + mpn7 + mpn10
  F3 =~ mpn3 + mpn5 + mpn8
  F3 ~ F1
  F2 ~ F3"
fit <- sem(m4, d, likelihood = "wishart")
summary(fit, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

```{r, tidy = FALSE}
m5 <- "
  F1 =~ mpn2 + mpn4 + mpn6 + mpn9
  F3 =~ mpn3 + mpn5 + mpn8
  F3 ~ F1
  mpn7 ~ F3"
fit <- sem(m5, d, likelihood = "wishart")
summary(fit, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

```
p = 36
q = 5 loadings + 7 measurement errors + 2 causal paths + 1 exo variance + 2 disturbances of endogenous
df = 19
```

Suppose that path from F2 to mpn7 is fixed to 1. Then disturbance of F2 and measurement error of mpn7 are the same thing.

```{r, tidy = FALSE}
m6 <- "
  F1 =~ mpn2 + mpn4 + mpn6 + mpn9
  F3 =~ mpn3 + mpn5 + mpn8
  F2 =~ mpn7
  F3 ~ F1
  F2 ~ F3
  mpn7 ~~ 11.42*mpn7
"
fit6 <- sem(m6, d, likelihood = "wishart")
summary(fit6, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
semPaths(fit6, layout = "tree", whatLabels = "est")
```


***

```{r}
sessionInfo()
```

