11/21/2013 - Mean Structures
===============================================================================

```{r, message = FALSE, warning = FALSE}
library(lavaan)
library(semPlot)
library(lmSupport)
d <- lm.readDat("../data/data_apgar.dat")
```

### How do you find the mean of X without using `mean`?

```{r}
d$one <- 1
m1 <- lm(data = d, smokes ~ 0 + one)
summary(m1)
```

The coefficient for `one` is the mean. This is a one-sample t-test too.

```{r}
m2 <- lm(data = d, smokes ~ 1)
summary(m2)
```

The path from the exogenous variable `one` to `smokes` is the mean.

Now, how to find the mean of an endogenous variable?

```{r}
m3 <- lm(data = d, gestat ~ 0 + smokes + one)
summary(m3)
```

The mean of `gestat` is the total effect of `one` on `gestat`.

```{r}
# Indirect effect + direct effect
unname(coef(m3)["smokes"] * coef(m1)["one"] + coef(m3)["one"])
mean(d$gestat)
```

## Mean structures

Recall `apgar2`.

```{r, tidy=FALSE}
apgar2 <- "
  apgar ~ gestat + wgtgain
  gestat ~ smokes
"
fit1 <- sem(apgar2, d, likelihood = "wishart")
summary(fit1)
```

Including the means structure adds 4 new pieces of information.

```{r}
(p <- 4 * (4 + 3) / 2)
(q <- 3 + 1 + 4 + 4)
p - q
```

`smokes` and `wgtgain` are no longer exogenous variables so we have to tell lavaan to let the two variables covary with `~~`. 

```{r, tidy=FALSE}
apgar2ms <- "
  apgar ~ gestat + wgtgain
  gestat ~ smokes
  apgar ~ 1
  gestat ~ 1
  wgtgain ~ 1
  smokes ~ 1
  wgtgain ~~ smokes
"
fit1ms <- sem(apgar2ms, d, likelihood = "wishart")
summary(fit1ms)
semPaths(fit1ms, whatLabels = "est")
```

Set `meanstructure = TRUE` instead of adding intercept terms.

```{r, tidy=FALSE}
fit1ms2 <- sem(apgar2, d, likelihood = "wishart", meanstructure = TRUE)
summary(fit1ms2)
```

The four means are replicated perfectly by the total effects for each variable. We use all four means in the model, so we have a "just-identified mean structure", so the means are perfectly replicated. The chi-square does not change with the addition of the (just-identified) mean structure. 

### Under-identified mean structure

The following model has 5 intercept paths but only 4 means. The mean structure is under-identified.

```{r, tidy=FALSE}
m_ex <- "
  F1 =~ v1 + v2 + v3 + v4
  F1 ~ 1
  v1 ~ 1
  v2 ~ 1
  v3 ~ 1
  v4 ~ 1
"
```


### Latent growth models

The test whether Change ~ 1 is different from 0 is all we care about in this kind of model.

```{r, tidy=FALSE}
m_latent_growth <- "
  Initial =~ 1*t1 + 1*t2 + 1*t3 + 1*t4
  Change  =~ 0*t1 + 1*t2 + 2*t3 + 3*t4
  Change ~~ Initial
  Initial ~ 1
  Change ~ 1
"
# cov: p = 14, q = 5, df_cov = 5
# mean: p = 4, q = 2, df_mean = 2
```




***

```{r}
sessionInfo()
```

