Structural Regression Models
===============================================================================

```{r, warning = FALSE, message = FALSE}
library(lavaan)
library(foreign)
data_carnet <- read.spss("../data/data_carnet3.sav", to.data.frame = TRUE)
round(psych::describe(data_carnet, skew = FALSE), 2)
```

> Please find below the R script I used in class today. In models 6 and 7 (which are mathematically equivalent), we are assuming that mpn7, the single indicator of the factor F2, has been measured with perfect reliability. In model 8, we are assuming that the reliability of mpn7 is .7, i.e., that 30% of its variability is due to random noise.


The Measurement Model
-------------------------------------------------------------------------------


### One latent variable

```{r, tidy = FALSE}
m1 <- '
  # latent variable definitions
    Fsingle =~ mpn1 + mpn2 + mpn3 + mpn4 + mpn5 + 
               mpn6 + mpn7 + mpn8 + mpn9 + mpn10'
fit1 <- sem(m1, data_carnet, likelihood = "wishart")
summary(fit1, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
residuals(fit1, type = "cor")
residuals(fit1, type = "standardized")
```

### Three latent variables

```{r, tidy = FALSE}
m2 <- '
  # latent variable definitions
    F1 =~ mpn2 + mpn4 + mpn6 + mpn9
    F2 =~ mpn1 + mpn7 + mpn10
    F3 =~ mpn3 + mpn5 + mpn8'
fit2 <- sem(m2, data_carnet, likelihood = "wishart")
summary(fit2, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
residuals(fit2, type = "cor")
residuals(fit2, type = "standardized")
```

Problem: Factors F1 and F3 are too highly correlated (r = .94).

### Two latent variables

```{r, tidy = FALSE}
m3 <- '
  # latent variable definitions
    F1 =~ mpn2 + mpn3 + mpn4 + mpn5 + mpn6 + mpn8 + mpn9
    F2 =~ mpn1 + mpn7 + mpn10'
fit3 <- sem(m3, data_carnet, likelihood = "wishart")
summary(fit3, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
residuals(fit3, type = "cor")
residuals(fit3, type = "standardized")
anova(fit1, fit3)
```

## Introduction to structural models

Let's ignore the fact that F1 and F3 are too highly correlated.

### Our first structural model: three causal paths

```{r, tidy = FALSE}
m4 <- '
  # latent variable definitions
    F1 =~ mpn2 + mpn4 + mpn6 + mpn9
    F2 =~ mpn1 + mpn7 + mpn10
    F3 =~ mpn3 + mpn5 + mpn8
  # regressions
    F3 ~ F1
    F2 ~ F3 + F1'
fit4 <- sem(m4, data_carnet, likelihood = "wishart")
summary(fit4, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
anova(fit2,fit4)
```

### Two causal paths (over-identified structural component)

```{r, tidy = FALSE}
m5 <- '
  # latent variable definitions
    F1 =~ mpn2 + mpn4 + mpn6 + mpn9
    F2 =~ mpn1 + mpn7 + mpn10
    F3 =~ mpn3 + mpn5 + mpn8
  # regressions
    F3 ~ F1
    F2 ~ F3'
fit5 <- sem(m5, data_carnet, likelihood = "wishart")
summary(fit5, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

### Model where outcome variable is an observed (manifest) variable

```{r, tidy = FALSE}
m6 <- '
  # latent variable definitions
    F1 =~ mpn2 + mpn4 + mpn6 + mpn9
    F3 =~ mpn3 + mpn5 + mpn8
  # regressions
    F3 ~ F1
    mpn7 ~ F3'
fit6 <- sem(m6, data_carnet, likelihood = "wishart")
summary(fit6, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

### One factor has only one indicator (same model, written differently)

Note: in both models, which are equivalent, we are assuming that the latent factor explains 100% of the variance in its (single) indicator. Probably an unreasonable assumption.

```{r, tidy = FALSE}
m7 <- '
  # latent variable definitions
    F1 =~ mpn2 + mpn4 + mpn6 + mpn9
    F2 =~ mpn7
    F3 =~ mpn3 + mpn5 + mpn8
  # regressions
    F3 ~ F1
    F2 ~ F3
  # residual correlations
    mpn7 ~~ 0*mpn7'
fit7 <- sem(m7, data_carnet, likelihood = "wishart")
summary(fit7, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
# Get the sample covariance matrix
inspect(fit7, "sampstat")
```

### Fixed measurement error

Our third structural model: same model, but this time we are fixing the variance of the measurement error of mpn7 to 30% of the variance of mpn7.

```{r, tidy = FALSE}
m8 <- '
  # latent variable definitions
    F1 =~ mpn2 + mpn4 + mpn6 + mpn9
    F2 =~ mpn7
    F3 =~ mpn3 + mpn5 + mpn8
  # regressions
    F3 ~ F1
    F2 ~ F3
  # residual correlations
    mpn7 ~~ 11.42*mpn7'
fit8 <- sem(m8, data_carnet, likelihood = "wishart")
summary(fit8, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
```

***

```{r}
sessionInfo()
```
