```{r, echo = FALSE, message = FALSE}
library(lavaan)
```


Mean structures
===============================================================================

```{r}
d <- read.table(file="../data/data_apgar.dat", header = TRUE, row.names = "SubID")
psych::describe(d)
d$one <- 1
```

## Use regression analysis to get the mean

### Get the mean of X

The coefficient for `one` is the mean of X:

```{r}
reg1 <- lm(smokes ~ 0 + one, data = d)
summary(reg1)
```

Using the intercept term instead of regressing onto `one` yields an equivalent model.

```{r}
reg1b <- lm(smokes ~ 1, data = d)
```

### Get the mean of Y

The mean of Y is the total effect of `one` on Y:

```{r}
reg2 <- lm(gestat ~ 0 + smokes + one, data = d)
summary(reg2)
# Indirect effect
one_to_smokes <- reg1$coefficients[["one"]]
smokes_to_gestat <- reg2$coefficients[["smokes"]]
# Direct effect
one_to_gestat <- reg2$coefficients[["one"]]
# Compute mean
one_to_smokes * smokes_to_gestat + one_to_gestat
mean(d$gestat)
```

As above, we get equivalent models by using an intercept term instead of regressing onto `one`.

```{r}
reg2b <- lm(gestat ~ 1 + smokes, data = d)
# Don't even need the "~ 1" part if there is another predictor.
reg2c <- lm(gestat ~ smokes, data = d)
```

## Using mean structures in `lavaan`

```{r, tidy = FALSE}
m2 <- '
  gestat ~ smokes
  apgar ~ wgtgain
  apgar ~ gestat'
fit2 <- sem(m2, data = d, likelihood = "wishart")
summary(fit2, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
residuals(fit2, type = "cor")
residuals(fit2, type = "standardized")
```

Model `apgar11` tries to add a mean-structure to `apgar2`. `lavaan` does not automatically estimate the covariance between the exogenous variables, so have to manually specify it, as in model `apgar12`.

```{r, tidy = FALSE}
m11 <- '
  gestat ~ smokes
  apgar ~ wgtgain
  apgar ~ gestat
  smokes + wgtgain + gestat + apgar ~ 1'
fit11 <- sem(m11, data = d, likelihood = "wishart")
summary(fit11, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
residuals(fit11, type = "cor")
residuals(fit11, type = "standardized")
```

Specify the covariance between the exogenous variables.

```{r, tidy = FALSE}
m12 <- '
  gestat ~ smokes
  apgar ~ wgtgain
  apgar ~ gestat
  smokes ~~ wgtgain
  smokes + wgtgain + gestat + apgar ~ 1'
fit12 <- sem(m12, data = d, likelihood = "wishart")
summary(fit12, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
residuals(fit12, type = "cor")
residuals(fit12, type = "standardized")
```

We can also use `meanstructure = TRUE` instead of specifying all the intercept terms.

```{r, tidy = FALSE}
m13 <- '
  gestat ~ smokes
  apgar ~ wgtgain
  apgar ~ gestat
  smokes ~~ wgtgain'
fit13 <- sem(m13, data = d, likelihood = "wishart", meanstructure = TRUE)
summary(fit13, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
residuals(fit13, type = "cor")
residuals(fit13, type = "standardized")
```

***

```{r}
sessionInfo()
```

