Questions about Rex Kline's book (for Tues., Nov. 26)
===============================================================================

> Read chapter 11 of Kline's book be prepared to answer the questions below. Please also do the data analysis exercise described at the end.

```{r, message = FALSE, warning=FALSE}
# Preliminary steps
library(foreign)
library(lavaan)
library(semPlot)

# Make a function to print out the constrained/free parameters from a model.
inspect_parameters <- function(model) {
  param_table <- inspect(model, what = "list")
  # Simplify the table
  param_table <- transform(param_table, call = paste(lhs, op, rhs, sep = " "))
  param_table <- param_table[c("call", "free", "group", "user", "ustart")]
  # Sort by the `free` column
  param_table[order(param_table$free, param_table$call), ]
}

# Load the data-sets
d_intell_fem <- read.spss("../data/Grnt_fem.sav", to.data.frame = TRUE)
d_intell_mal <- read.spss("../data/Grnt_mal.sav", to.data.frame = TRUE)
d_intell_fem$gender <- "female"
d_intell_mal$gender <- "male"
d <- rbind(d_intell_fem, d_intell_mal)
# Make a numeric version of `gender`
d <- transform(d, sex = ifelse(gender == "female", 0, 1))

# Data for exercises 15-16
covmatrix <- matrix(
  c(6.502500, 4.793490, 4.652985, 4.373760, 0.199410,
    4.793490, 5.808100, 4.521401, 4.391984, 0.221720,
    4.652985, 4.521401, 6.604900, 5.165700, 0.242351,
    4.373760, 4.391984, 5.165700, 7.182400, 0.295872,
    0.199410, 0.221720, 0.242351, 0.295872, 0.052900),
  nrow = 5, dimnames = list(
    c("sales1", "sales2", "sales3", "sales4", "region"),
    c("sales1", "sales2", "sales3", "sales4", "region")))
myDataMeans <- c(6.08, 7.22, 8.14, 9.38, 0.48)
names(myDataMeans) <- c("sales1", "sales2", "sales3", "sales4", "region")
```


#### 1. Assuming you include a mean structure, how do you determine the model-implied (predicted) mean for a given variable?

The model-implied mean of X is the total effect of the constant variable ("1") on X. For an endogenous variable, this is the direct effect ("the intercept") plus the indirect effects (e.g., the mean of the exogenous variable Y ~ 1 times the path coefficient X ~ Y.)



#### 2. What is the mean of a variable assumed to be if it is excluded from the mean structure?

Zero.



#### 3. "A heavily over-identified covariance structure can compensate for a mildly under-identified mean structure, and vice versa." True or false?

False. "The identification status of a mean structure must be considered separately from that of the covariance structure" (p. 303).



#### 4. What are the implications of adding a just-identified mean structure for model fit? 

> In other words: If you add a just-identified mean structure, does this usually lead to A) better, B) worse, or C) equal model fit compared to the same model without mean structure?

The just-identified mean structure will precisely estimate the means of the observed variables. The covariance structure will not be affected by the added mean structure, so the model will have C) equal model fit compared to the same model without the mean structure. 



#### 5. Compare the advantages and disadvantages of HLM and SEM for the estimation of latent growth models.

+ SEM growth models require **time-structured data**. This means that the subjects are measured with uniform inter-measurement intervals. HLM does not require time-structured data.
+ HLM handles missing data and unbalanced data more flexibly than SEM.
+ SEM provides indices for evaluating whole-model fit.
+ SEM can handle multiple growth curves (multiple outcomes) simultaneously in a single model.
+ SEM can model factors as outcome variables. It can use factors as predictor variables as well, whereas HLM cannot, and it is generally easier to compute indirect effects among these latent factors.



#### 6. When we specify a latent growth model in SEM, do we usually allow the "Initial Status" factor and the "Linear Change" factor to covary? If not, why not? If yes, what is the meaning of this covariance?  

Yes. The covariance describes how much the Initial Status predicts later rates of Linear Change. It makes sense that the baseline level would be correlated with later changes away from that baseline level.



#### 7. Compare latent growth analysis in SEM with more traditional data analysis techniques, such as "repeated measures analysis of variance" (within-subject ANOVA), "multivariate analysis of variance" (MANOVA). What assumptions do these different techniques make for the measurement errors? How do they treat individual differences in growth trajectories?

SEM provides a mechanism for modeling measurement error, and we can allow measurement errors to covary with each other. ANOVA assumes that error variance is constant and independent (tough assumptions for repeated measures within subjects). MANOVA requires less strict assumptions (allowing non-independent errors). "Both ANOVA and MANOVA treat individual differences in growth trajectories as error variance" (p. 307) whereas latent growth models can model such differences.



#### 8. What would we do, if we wanted to estimate a curvilinear trend in addition to the linear trend specified in Figure 11.2? What would be the df of the model that includes also a curvilinear trend?

Figure 11.2 has:

```
P = (4 * 7) / 2 = 14.
Q = 2 factor variances + 4 measurement errors variances + 1 factor covariance + 
    3 error covariances + 2 factor means
Q = 12 
df = 14 - 12 = 2
```

Adding the quadratic change factor would also add:

```
Q_new = Q + 1 factor mean + 1 factor variance + 2 factor covariances
Q_new = 16
df = 14 - 16 = -2
```

#### 9. Consider the "intell" data we analyzed several weeks ago. Be prepared to draw on the blackboard a multi-group measurement model that allows us to test for configural invariance.

> As a reminder, we tested a CFA model with two factors and three indicators per factor. The factor "spatial" was assumed to influence the indicators "visperc", cubes", and "lozenges", and the factor "verbal" was assumed to influence the indicators "paragraph", "sentence", and "wordmean". There were two samples, women ("Grnt_fem.sav") and men ("Grnt_mal.sav"). We already did that. Just redraw the model we drew several weeks ago. What are the # of observations, # of estimated parameters, and df of this model? Is this model at least just-identified?

```{r, warning = FALSE, tidy = FALSE}
# Configural invariance model
m_form <- '
  spatial =~ visperc + cubes + lozenges 
  verbal =~ paragrap + sentence + wordmean'
h_form <- cfa(m_form, data = d, likelihood = "wishart", group = "gender", 
              meanstructure = FALSE)
summary(h_form, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
semPaths(h_form, panelGroups = TRUE, ask = FALSE, title = FALSE)
```

The model is over-identified.

```
P = (6 * 7) / 2 * 2 samples = 42
Q_base = 2 factor variances + 1 factor covariance + 
         4 factor loadings + 6 measurement errors
Q = 13 * 2 samples = 26
df = 42 - 26 = 16
```



#### 10. Modify this model to test for "construct-level metric invariance". 

> If you don't remember what this is, reread chapter 9 of Kline's book. You will have to constrain certain parameters to be the same across groups. Do so by assigning to them the same lower-case letter in both samples. What are the # of observations, # of estimated parameters, and df of this model? Is this model at least just-identified?

Construct-level metric invariance specifies equal factor loadings across both groups. We have to estimate four fewer parameters because the same factor loadings will be used for each sample. Therefore, P = 42, Q = 26 - 4 = 22, df = 16 + 4 = 20. The model is overidentified.

```{r, tidy = FALSE}
m_lambda <- '
  spatial =~ visperc + c(a, a)*cubes + c(b, b)*lozenges 
  verbal =~ paragrap + c(c, c)*sentence + c(d, d)*wordmean'
h_lambda <- cfa(m_lambda, data = d, likelihood = "wishart", group = "gender", 
              meanstructure = FALSE)
inspect_parameters(h_lambda)
summary(h_lambda, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
# semPaths(h_lambda, panelGroups = TRUE, ask = FALSE, title = FALSE)
```

The chi-square difference test indicates that the fit is not significantly worse, so H-lambda is retained.

```{r}
anova(h_form, h_lambda)
```


#### 11. Add a mean structure to the previous model. In each sample, specify a path from the constant to each of the six indicators and to each of the two factors. 

```{r, tidy = FALSE}
# Write out the model syntax long-hand
m_lambda_1s <- '
  spatial =~ visperc + c(a, a)*cubes + c(b, b)*lozenges 
  verbal =~ paragrap + c(c, c)*sentence + c(d, d)*wordmean
  paragrap + sentence + wordmean + visperc + cubes + lozenges ~ 1
  spatial + verbal ~ 1'
# I think the under-indentified mean structure upsets lavaan
h_lambda_1s <- cfa(m_lambda_1s, data = d, likelihood = "wishart", group = "gender")
inspect_parameters(h_lambda_1s)
summary(h_lambda_1s, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
semPaths(h_lambda_1s, panelGroups = TRUE, ask = FALSE, title = FALSE)
```

> What are the # of observations, # of estimated parameters, and df of the covariance structure? Is the covariance structure at least just-identified? 

The covariance structure is unaffected. P = 42, Q = 22, df = 20. It is over-identified.

> What are the # of observations, # of estimated parameters, and df of the mean structure? Is the mean structure at least just-identified? 

The mean structure is under-identified.

```
P = 6 means * 2 samples = 12
Q = (6 indicators means + 2 factor means) * 2 samples = 16
df = -4
```

> What are the # of observations, # of estimated parameters, and df of the total model? Is the total model at least just-identified?

The total model is over-indentified.

```
P = 54
Q = 22 covariance parameters + 16 mean parameters = 38
df = 16
```




#### 12. Modify the model so that we can test whether there are gender differences on the two factors. 

> You'll find detailed instructions on the bottom half of page 317 and on the bottom half of page 318 in Kline's book. Make sure both the covariance structure and the mean structure are at least just-identified. Once you are done, determine the df for the entire model.  

First, we should test if we can use the same intercepts for the two groups. We do this taking our `h_lambda` model (with the equal loadings for the two groups) and adding on a mean structure. We compare a mean structure with freely estimated intercepts to one with equal intercepts for the two groups. 

```{r, tidy = FALSE}
m_lambda_tau <- '
  spatial =~ visperc + cubes + lozenges 
  verbal =~ paragrap + sentence + wordmean
  verbal ~ c(a, b)*1
  spatial ~ c(c, d)*1'
# Freely estimated intercepts
h_form_mean <- cfa(m_lambda_tau, data = d, likelihood = "wishart", 
                    group = "gender", group.equal = c("loadings"))
inspect_parameters(h_form_mean)
# semPaths(h_form_mean, panelGroups = TRUE, ask = FALSE, title = FALSE)
# Now we constrain the intercepts to be the same for each group
h_lambda_tau <- cfa(m_lambda_tau, data = d, likelihood = "wishart", 
                    group = "gender", group.equal = c("loadings", "intercepts"))
inspect_parameters(h_lambda_tau)
summary(h_lambda_tau, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
# semPaths(h_lambda_tau, panelGroups = TRUE, ask = FALSE, title = FALSE)
anova(h_form_mean, h_lambda_tau)
```

`h_lambda_tau` is retained, meaning that we can use the same intercepts and factor loadings for the two groups. 

Now we investigate whether there is a difference between the groups.

```{r, tidy = FALSE}
m_factor_means <- '
  spatial =~ visperc + cubes + lozenges 
  verbal =~ paragrap + sentence + wordmean
  verbal ~ c(a, b)*1
  spatial ~ c(c, d)*1
  a == 0
  c == 0'
h_factor_means <- cfa(m_factor_means, data = d, likelihood = "wishart", 
                      group = "gender", group.equal = c("loadings", "intercepts"))
inspect_parameters(h_factor_means)
summary(h_factor_means, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
# semPaths(h_factor_means, panelGroups = TRUE, ask = FALSE, title = FALSE)
```

The male subjects had a mean spatial factor 1.01 units higher than the female subjects, but this difference is not significant, z = 1.2, p = 0.23. The female subjects had a mean verbal factor 0.96 units higher than the male subjects, and this difference falls just short of achieving significance, z = 1.82, p = 0.068. 

```
Group 1 [female]:
Intercepts:        Estimate  Std.err  Z-value  P(>|z|)   Std.lv  Std.all
    verbal    (a)     0.000    0.000    7.677    0.000    0.000    0.000
    spatial   (c)     0.000    0.000    6.640    0.000    0.000    0.000
Group 2 [male]:
Intercepts:
    verbal    (b)    -0.956    0.524   -1.823    0.068   -0.358   -0.358
    spatial   (d)     1.066    0.887    1.201    0.230    0.267    0.267
```



```
P_mean = 6 * 2 samples = 12
Q_mean = 6 shared indicator intercepts + 2 estimated factor means
df_mean = 4

P_cov = 42
Q_cov = (6 measurement errors + 2 factor variances + 1 factor covariance) * 2 +
        (4 shared factor loadings) = 22
df_cov = 20

df_all = 24
```

#### 13. Respecify the model as a MIMIC model in which the two factors are regressed on a dichotomous cause indicator "gender".  What are the # of observations, # of estimated parameters, and df of this model? Is this model at least just-identified?

```{r, tidy = FALSE}
m_mimic <- '
  spatial =~ visperc + cubes + lozenges 
  verbal =~ paragrap + sentence + wordmean
  spatial ~ sex
  verbal ~ sex
  sex ~~ sex
  spatial ~~ verbal'
h_mimic <- cfa(m_mimic, data = d, likelihood = "wishart")
inspect_parameters(h_mimic)
summary(h_mimic, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
semPaths(h_mimic)
```

This model is over-identified.

```
P = 7 * 8 / 2 = 28
Q = 2 factor disturbances + 6 measurement errors + 4 factor loadings + 
    1 factor disturbance covariance + 2 paths + 1 exogenous variance
Q = 16
df = 12
```

These estimates are close to the estimated group differences in the previous model.

```
Regressions:       Estimate  Std.err  Z-value  P(>|z|)   Std.lv  Std.all
  spatial ~
    sex               1.086    0.858    1.266    0.205    0.254    0.127
  verbal ~
    sex              -0.949    0.518   -1.831    0.067   -0.321   -0.161
```

#### 14. A MIMIC model to test for factor differences across groups makes a certain assumption, and yet does not allow us to test for this assumption. What is this assumption?

Measurement invariance across the groups.



#### 15. The manager of Abercrombie & Fitch would like to test whether sales increased over a 4-year period. She looks at the yearly sales in 51 stores in two regions (West and East). 

> The covariance matrix and the means are below. Be prepared to draw the model and to show your computations for the df. Using lavaan, run a latent growth model according to the example in Figure 11.2 (but without correlations between the measurement errors). What conclusions can the manager draw: Is there a linear increase in sales?

```{r, tidy = FALSE}
m_growth <- "
  IS =~ 1*sales1 + 1*sales2 + 1*sales3 + 1*sales4
  LC =~ 0*sales1 + 1*sales2 + 2*sales3 + 3*sales4
  LC + IS ~ 1
  sales1 + sales2 + sales3 + sales4 ~ 0*1"
h_growth <- cfa(m_growth, sample.cov = covmatrix, sample.mean = myDataMeans, 
                sample.nobs = 51, likelihood = "wishart")
summary(h_growth, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
semPaths(h_growth)
```


```
P = 4 * 7 / 2 = 14
Q_cov  = 7 = 2 factor variances + 1 factor covariance + 4 measurement errors
Q_mean = 2 factor means
Q = 9
df = 5
```

Yes there is a significant linear effect of time on sales. On average, sales increase by 1.08 units annually, z = 10.53, p < 0.001.


#### 16. The manager also wants to know if there are differences between the two regions. 

> Were sales higher in one region than in the other at year 1? And did the sales in one region increase faster than in the other region? Using lavaan, run a latent growth model according to the example in Figure 11.3 (replace "gender" by "region" and drop "family status" from the model). Be prepared to draw the model and to show your computations for the df. What conclusions can the manager draw?


```{r, tidy = FALSE}
m_growth_2 <- "
  IS =~ 1*sales1 + 1*sales2 + 1*sales3 + 1*sales4
  LC =~ 0*sales1 + 1*sales2 + 2*sales3 + 3*sales4
  LC ~ region
  IS ~ region
  LC ~~ IS
  LC + IS + region ~ 1
  sales1 + sales2 + sales3 + sales4 ~ 0*1"
h_growth_2 <- cfa(m_growth_2, sample.cov = covmatrix, sample.mean = myDataMeans, 
                  sample.nobs = 51, likelihood = "wishart")
summary(h_growth_2, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE)
semPaths(h_growth_2)
```

```
P = 5(8) / 2 = 20
Q_cov  = 4 measurement errors + 2 paths + 2 factor disturbances + 
         1 exogenous variance + 1 disturbance covariance
Q_mean = 2 factor intercepts + 1 variable mean
Q = 13
df = 7
```

The path from Region to Initial Status is significant, meaning that one region had greater sales initially, b = 3.68, z = 2.59, p = 0.01. The path from Region is not significant, so the regions did not differ significantly in sales growth, b = 0.58, z = 1.32, p = 0.19.




***

```{r}
sessionInfo()
```

